Incorporating periodic variability in hidden Markov models for animal movement

### 2016-07-11 paper3.Rnw
lots of change, almost done

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1). The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components:

@@

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
fit of each model type.
(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
captures serial correlation at both short and long scales. 
We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
                                    , y=deltaBIC
                                    , colour=model
                                    , linetype=model
                                    , group=model))
  + facet_wrap(~mod,ncol = 2,scales = "free_x")
  + scale_size_continuous(name="# of states")
  + scale_x_continuous(breaks=seq(0,300,by=10))
  + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
  + scale_linetype_manual(values=linetypev)
  + scale_colour_brewer(palette="Dark2")
  + labs(y=expression(paste(Delta, "BIC"))
         , x="Number of States") 
  + geom_line()
  + zmargin)
@

>>>>

@@

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1). The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions. Then the number of free parameters of an HMM is:

K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall goodness of
fit of each model type.
(2) Comparing average step-length by hour of day for the observed data and for data simulated from the models shows how well a particular class of models can capture the diurnal variation in behaviour.
(3) Comparing temporal autocorrelations for the observed data and for data simulated from the models shows how well a particular class of models
captures serial correlation at both short and long scales.

\subsection{BICS}
Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity (Supplementary material, Table 1). The number of free parameters in an HMM can be generalize by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{I}, k_{t}, k,{e}$ be the parameters of the prior distribution, transition function and emission distributions. Then the number of free parameters of an HMM is:

K = $$ k_{i} * (n-1) + k_{t}*n*(n-1) + k_{e}*n $$ ...

\subsection{Comparing observed data against simulation and Viterbi predictions}

We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex.
\todo{need to note here that using the Viterbi algorithm to simulate double-counts the data ...}

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Overall adjusted BIC comparisons among models, cat \\#1. \\textbf{todo: use dashed lines for FMMs (leadup to next figure)?}",fig.width=8>>=
#L <- load("cat1sumdat.RData")
sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
                    mod = ifelse(grepl("HMM",model),"no","yes")
)
linetypev <- c(rep(2,2),rep(1,5))
(adj_BIC_plot <- ggplot(sumdat, aes(x=nstates
                                    , y=deltaBIC
                                    , colour=model
                                    , linetype=model))
+ facet_wrap(~ type,ncol=4)
+ scale_size_continuous( name="# of parameters")
+ scale_x_continuous(expand=c(0,0.5),breaks=2:7)
+ geom_point(aes(size=parameters))
+ scale_linetype_manual(values=linetypev)
+ scale_colour_brewer(palette="Dark2")
+ labs(y=expression(paste(Delta, "BIC"))
       , x="Number of States") 
+ geom_line() 
+ zmargin)

### I want this picture but it is too big or something

# sumdat <- transform(readRDS("summary_stats/cat.1.df.RDS")$HMMsumdf,
#                     mod = ifelse(type=="HMM + TH","HMM + TH","FMMs and HMM")
# )
# linetypev <- c(rep(2,2),rep(1,5))
# (adj_BIC_plot <- ggplot(sumdat, aes(x=parameters
#                                     , y=deltaBIC
#                                     , colour=model
#                                     , linetype=model
#                                     , group=model))
#   + facet_wrap(~mod,ncol = 2,scales = "free_x")
#   + scale_size_continuous(name="# of states")
#   + scale_x_continuous(breaks=seq(0,300,by=10))
#   + geom_point(aes(shape=factor(nstates),size=1.5,alpha=nstates))
#   + scale_linetype_manual(values=linetypev)
#   + scale_colour_brewer(palette="Dark2")
#   + labs(y=expression(paste(Delta, "BIC"))
#          , x="Number of States") 
#   + geom_line()
#   + zmargin)
@
